Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2020-11-26 manuscript_Huang.tex
Updated by Reik.

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_fBm.pdf}
\caption{\small{(Color online) Same as in Fig.~\ref{fig:CElogistic}, but for fractional Brownian motion with a Hurst exponent $H \in (0, 1)$ varying with a step size of $\Delta H=0.05$. Note that the normalized entropy is generally decreasing with increasing $H$. {\color{red} Should the text arrows are confusing while indicating the increasing direction of $H$? }}  \label{fig:CEfBm}}
\end{figure*}

@@

While we have previously demonstrated the suitability of OPTN based SCMs for tracing changes in dynamical complexity for time series from deterministic dynamical systems, in the following we will use two different experimental time series to demonstrate that SCMs also successfully characterize the complexity properties of real-world systems, showing distinct values for different types of dynamics.

The first example originates from a laboratory fluid experiment of baroclinic instability which is used to study patterns of cyclones and anticyclones in the Earth's atmosphere~\cite{Read_jfm_1992,ZouEPJST2008}. Depending on the parameters of rotation rate, temperature difference, viscosity and fluid density, this experimental system exhibits a rich variety of flow regimes. In particular, we focus on the following regimes: 
\begin{enumerate}[(i)]
\item {Stable fluid:} The temperature signal of the stable flow exhibits periodic oscillations as the wave drift patterns arrive at the fixed point of measurements. 
\item {Quasi-periodic 2-frequency amplitude vacillation:} This case is identified as a 2-frequency quasi-periodic amplitude flux. The wave drift is composed of slow and regular oscillations in the temperature signal, but a fast modulation in the amplitude is also visible. AV-2 is characterized by a periodic growth and drop in wave amplitude with little change in waveform. 
\item {Quasi-periodic 3-frequency amplitude vacillation:} We also analyze a quasi-periodic 3-frequency amplitude vacillation time series which is obtained from a three-dimensional direct numerical simulation in an air-filled rotating baroclinic instability experiment~\cite{Read_jfm_1992}. 
\item {Modulated amplitude vacillation:} This case is identified as a low-dimensional stream, chaotically modulated and with waves of varying amplitudes. Amplitude modulation results in a complex temperature dynamics.
\end{enumerate}

@@

In both sets of time series, the OPTN based SCMs enhance our understanding of the underlying nonlinear system, which can be seen from the results summarized in Fig.~\ref{fig:fluid}. For the fluid data (Fig.~\ref{fig:fluid}(a)), $\mathcal{H}_O$, $\mathcal{H}_W$, $\mathcal{C}_W$, $\mathcal{H}_E$ and $\mathcal{C}_E$ all show consistent changes of the complexity and entropy values between stable wave solution (SW), quasi-periodic 2-frequency amplitude vacillation (AV-2), quasi-periodic 3-frequency amplitude vacillation (AV-3) and chaotic modulated amplitude vacillation (MAV). Specifically, the complexity of these four cases has the following order: SW has the lowest complexity since it shows periodic oscillations with trivial recurrence, while MAV has the highest complexity since it presents chaotically modulated amplitude vacillation. Quasiperiodic states exhibit intermediate complexity values since these two cases have non-trivial recurrences while still featuring a certain degree of regularity. Interestingly, AV-3 presents higher complexity than AV-2. The order of complexity values between periodic, quasiperiodic and chaotic solutions still holds for $\mathcal{C}_O$, but the corresponding values become very similar. In particular, it becomes hard for $\mathcal{C}_O$ to identify the difference between AV-2 and AV-3.

>>>>

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_fBm.pdf}
\caption{\small{(Color online) Same as in Fig.~\ref{fig:CElogistic}, but for fractional Brownian motion with a Hurst exponent $H \in (0, 1)$ varying with a step size of $\Delta H=0.05$. Note that the normalized entropy is generally decreasing with increasing $H$ (red arrow).}  \label{fig:CEfBm}}
\end{figure*}

@@

While we have previously demonstrated the suitability of OPTN based SCMs for tracing changes in dynamical complexity for time series from deterministic dynamical systems, in the following we will use two different sets of experimental time series to demonstrate that SCMs also successfully characterize the complexity properties of real-world systems, showing distinct values for different types of dynamics.

The first example originates from a laboratory fluid experiment of baroclinic instability which is used to study patterns of cyclones and anticyclones in the Earth's atmosphere~\cite{Read_jfm_1992,ZouEPJST2008}. Depending on the parameters of rotation rate, temperature difference, viscosity and fluid density, this experimental system exhibits a rich variety of flow regimes. In particular, we focus on the following regimes: 
\begin{enumerate}[(i)]
\item {Stable fluid:} The temperature signal of the stable flow exhibits periodic oscillations as the wave drift patterns arrive at the fixed point of measurements. 
\item {Quasi-periodic 2-frequency amplitude vacillation:} This case is identified as a 2-frequency quasi-periodic amplitude flux. The wave drift is composed of slow and regular oscillations in the temperature signal, but a fast modulation in the amplitude is also visible. AV-2 is characterized by a periodic growth and drop in wave amplitude with little change in waveform. 
\item {Quasi-periodic 3-frequency amplitude vacillation:} For the sake of completeness, we also analyze a quasi-periodic 3-frequency amplitude vacillation time series. Unlike for the other studied cases, this time series has not been obtained from experimental results, but from a three-dimensional direct numerical simulation of the air-filled rotating baroclinic instability experiment~\cite{Read_jfm_1992}. 
\item {Modulated amplitude vacillation:} This case is identified as a low-dimensional stream, chaotically modulated and with waves of varying amplitudes. Amplitude modulation results in a complex temperature dynamics.
\end{enumerate}

@@

In both sets of time series, the OPTN based SCMs enhance our understanding of the underlying nonlinear system, which can be seen from the results summarized in Fig.~\ref{fig:fluid}. For the fluid data (Fig.~\ref{fig:fluid}(a)), $\mathcal{H}_O$, $\mathcal{H}_W$, $\mathcal{C}_W$, $\mathcal{H}_E$ and $\mathcal{C}_E$ all show consistent changes of the complexity and entropy values between stable wave solution (SW), quasi-periodic 2-frequency amplitude vacillation (AV-2), quasi-periodic 3-frequency amplitude vacillation (AV-3) and chaotic modulated amplitude vacillation (MAV). Specifically, the complexity of these four cases has the following order: SW has the lowest complexity since it shows periodic oscillations with trivial recurrence, while MAV has the highest complexity since it presents chaotically modulated amplitude vacillation. Quasiperiodic states exhibit intermediate complexity values since these two cases have non-trivial recurrences while still featuring a certain degree of regularity. In our case, AV-3 presents a higher complexity than AV-2, which should however be interpreted with care since the AV-2 series originates from an experiment, while the AV-3 series has been taken from a numerical simulation of the same system (hence, there may be a certain level of observational noise present in AV-2 but not in AV-3). The order of complexity values between periodic, quasiperiodic and chaotic solutions still holds for $\mathcal{C}_O$, but the corresponding values become very similar. In particular, it becomes hard for $\mathcal{C}_O$ to identify the difference between AV-2 and AV-3.
